
\chapter*{ABSTRACT}

\footnotesize{
\begin{description}
  \item[TITLE:] \MakeUppercase{Detection of Diabetic Retinopathy in Fundus Images Using
  Deep Learning Algorithms.}
  \astfootnote{Bachelor Thesis}
  \item[AUTHOR:] {\autora}, {\autorb}
  \asttfootnote{\facultad. \escuela. Advisor: \director. Co-advisor: \codirectora}
  \item[KEYWORDS:] DIABETIC RETINOPATHY, CONVOLUTIONAL NEURAL NETWORKS, DEEP LEARNING,
  FUNDUS IMAGES, DIGITAL IMAGE PROCESSING, TRANSFER LEARNING.
  \item[DESCRIPTION:]\hfill \\ Diabetic retinopathy (DR) is a microvascular complication
  caused by damage to the blood vessels of the retina due to elevated blood sugar levels,
  common in people with diabetes and considered one of the leading causes of blindness
  worldwide. This disease is typically diagnosed through eye exams that assess the fundus.
  However, manual diagnosis is challenging because it is costly and time-consuming,
  limiting access in areas with scarce medical resources. This makes it necessary to
  seek automated methods that enable faster and more accurate disease detection. This
  work develops models based on convolutional neural networks (CNNs) to detect the
  disease and classify its severity levels using binary classifiers (presence/absence)
  and multiclass classifiers (five levels). Transfer learning was applied with the
  MobileNetV2, ResNet50V2, InceptionV3, and VGG19 architectures, trained on 35,121
  images from the public Kaggle dataset. Preprocessing included color normalization,
  peripheral masking, class balancing via subsampling and data augmentation,
  incorporating Multi-Head Attention blocks, GaussianNoise layers, and regularization
  techniques (Dropout, L2, early stopping). Models were evaluated using sensitivity,
  precision, F1-score, and AUC-ROC. MobileNetV2 was the best-performing model,
  achieving a sensitivity of 96.5\% in binary classification with confidence intervals
  (CI95\%: AUC-ROC 0.9636--0.99). For multiclass classification, a sensitivity of 65\%
  was achieved, identifying the green channel as the most effective due to its contrast
  for hemorrhagic lesions. External validation on the MESSIDOR-2, IDRiD, and FOSCAL
  datasets confirmed the model's generalization capability, demonstrating its viability
  as a clinical screening support tool for diabetic retinopathy.

\end{description}}\normalsize
% ------------------------------------------------------------------------
